\chapter*{Acknowledgments}

I truly feel fortunate to have the opportunities I did in the department of Bioinformatics under the guidance of Prof. Dr. Martin Hofmann-Apitius. Thank you for all the support you've given me over the past few years, both academic and personal. I would also like to express my gratitude to Prof. Dr. Diana Imhof for acceding to be an additional reviewer of my thesis as well as to the members of my defence committee.

To Dr. Daniel Domingo-Fernández, my academic advisor and scientific mentor, I really can’t seem to find the right words to say thank you. You made all of this possible. For the wealth of knowledge and all the guidance you’ve given me the past five years and for the scientist you’ve shaped me into, I owe you my sincerest gratitude. Your foresight and determination were the driving force for this work and you taught me such a great deal about perseverance. You taught me technical skills when I had none and to see the big picture when I was stuck in the details. Your expertise and your cleverness have taken me by surprise on many an occasion (although admittedly, my surprise could only ever be attributed to your humility) and helped me to navigate through challenges which I had thought were impassable. Ultimately, I know that you’re a wonderful scientist and a brilliant academic, but more than that, I know that you're an even better person who follows his dreams to the fullest and encourages each person he encounters to do the same. Thanks coach!

Special thanks are also in order for all of my fellow PhD students and colleagues at SCAI. Though the duration of my PhD largely coincided with the COVID-19 pandemic and our time together was ever so short, I am so grateful for all the wonderful discussions and engagement that still took place. In particular, to Rebeca Queiroz Figueiredo, Vinay Srinivas Bharadhwaj and Sepehr Golriz Khatami, it was such a pleasure working with you! To Alpha Tom Kodamullil, I am so glad to have had your support all these years since I first started at SCAI over six years ago. A heartfelt thank you to Meike Knieps and Alina Enns for always keeping things running so smoothly and for all of your patience. Warm regards to the Women in Science! It was my good fortune to have had the constant support of such talented and kind colleagues like you!

Finally, to my family and friends who are never so far, cheering me on and helping me to keep things in perspective, I am so thankful. To my mom and dad, my appreciation and gratitude towards you are boundless. You instilled in me the importance of integrity and a commitment to always seek knowledge, lessons that have stayed with me throughout my academic journey.



