\chapter*{Publications}

\section*{Thesis Publications}

\textsuperscript{\textdagger} Joint last authors

\begin{enumerate}

    \item \textbf{Mubeen S.}, Kodamullil A.T., Hofmann-Apitius M., and Domingo-Fern\'{a}ndez D. (2022). On the influence of several factors on pathway enrichment analysis. \textit{Briefings in Bioinformatics}, 23(3).
    \item \textbf{Mubeen S.}, Hoyt C. T., Gemünd A., Hofmann-Apitius M., Fröhlich H., and Domingo-Fern\'{a}ndez D. (2019). The Impact of Pathway Database Choice on Statistical Enrichment Analysis and Predictive Modeling. \textit{Frontiers in Genetics}, 10:1203.
    \item \textbf{Mubeen S.}, Bharadhwaj S. V., Kodamullil A.T., Gadiya Y., Hofmann-Apitius M., and Domingo-Fern\'{a}ndez D. (2021). DecoPath: A Web Application for Decoding Pathway Enrichment Analysis.  \textit{NAR Genomics and Bioinformatics}, 3(3): lqab087.
    \item Figueiredo R.Q., Raschka T., Kodamullil A.T., Hofmann-Apitius M., \textbf{Mubeen S.\textsuperscript{\textdagger}}., and Domingo-Fern\'{a}ndez D.\textsuperscript{\textdagger} (2021). Towards a global investigation of transcriptomic signatures through co-expression networks and pathway knowledge for the identification of disease mechanisms. \textit{Nucleic acid research}, 49(14): 7939–7953.
    \item Figueiredo R.Q., Díaz del Ser S., Raschka T., Hofmann-Apitius M., Kodamullil A.T., \textbf{Mubeen S.\textsuperscript{\textdagger}}., and Domingo-Fern\'{a}ndez D.\textsuperscript{\textdagger} (2022). Elucidating gene expression patterns across multiple biological contexts through a large-scale investigation of transcriptomic datasets. \textit{BMC Bioinformatics}, 23(1).
    \item Marín-Llaó J., \textbf{Mubeen S.}, Perera-Lluna A., Hofmann-Apitius M., Picart-Armada S.\textsuperscript{\textdagger}, and Domingo-Fern\'{a}ndez D.\textsuperscript{\textdagger} (2021). MultiPaths: a Python framework for analyzing biological networks using diffusion algorithms. \textit{Bioinformatics}, 37(1): 137-139.
    \item Rivas-Barragan D., \textbf{Mubeen S.}, Guim Bernat F., Hofmann-Apitius M., and Domingo-Fern\'{a}ndez D. (2020). Drug2ways: Reasoning over causal paths in biological networks for drug discovery. \textit{PLOS Computational Biology}, 16(12): e1008464.

\end{enumerate}

\section*{Other Publications}

\begin{enumerate}
    \setcounter{enumi}{7}
    \item Domingo-Fern\'{a}ndez D., \textbf{Mubeen S}., Marín-Llaó J, Hoyt C. T., and Hofmann-Apitius M. (2019). PathMe: Merging and exploring mechanistic pathway knowledge. \textit{BMC Bioinformatics}, 20(1): 243.
    \item Hoyt C. T., Domingo-Fern\'{a}ndez D., \textbf{Mubeen S.}, Marín-Llaó J, Konotopez A., Ebeling C., Birkenbihl C., English B., Müller S., Ali M., Colby S., Türei D., Palacio N. and Hofmann-Apitius M. (2019). Integration of Structured Knowledge Sources with Biological Expression Language. \textit{bioRxiv}, 631812.
    \item Golriz Khatami S., \textbf{Mubeen S.}, and Hofmann-Apitius, M. (2020). Data science in neurodegenerative disease: Its capabilities, limitations, and perspectives. \textit{Current opinion in neurology}, 33(2): 249.
    \item Golriz Khatami S., \textbf{Mubeen S.}, Bharadhwaj S. V., Kodamullil A.T., Hofmann-Apitius M., and Domingo-Fern\'{a}ndez D. (2020). Using predictive machine learning models for drug response simulation by calibrating patient-specific pathway signatures. \textit{npj Systems Biology and Applications}, 7(1): 40.
    \item Bharadhwaj S. V., Mehdi C. T., Birkenbihl C., \textbf{Mubeen S.}, Lehmann J., Hofmann-Apitius M., Hoyt C. T., and Domingo-Fern\'{a}ndez D. (2021). CLEP: A Hybrid Data- and Knowledge-Driven Framework for Generating Patient Representations. \textit{Bioinformatics}, 37(19): 3311-3318.
    \item Golriz Khatami S., Russo M. F., Domingo-Fern\'{a}ndez D., Zaliani A., \textbf{Mubeen S.}, Gadiya Y., … Hofmann-Apitius M. (2021). Curating, collecting, and cataloguing global COVID-19 datasets for the aim of predicting personalized risk. \textit{medRxiv}, 2021.11.14.21265797.
    \item Golriz Khatami S., Domingo-Fern\'{a}ndez D., \textbf{Mubeen S.}, Hoyt C. T., Robinson C., Karki R., … Hofmann-Apitius M. (2021). A Systems Biology Approach for Hypothesizing the Effect of Genetic Variants on Neuroimaging Features in Alzheimer’s Disease. \textit{Journal of Alzheimer's Disease}, 80(2): 831-840.
    \item Domingo-Fern\'{a}ndez D., Gadiya Y., Patel A., \textbf{Mubeen S.}, Rivas-Barragan D., Diana C., … Colluru V. (2021). Causal reasoning over knowledge graphs leveraging drug-perturbed and disease-specific transcriptomic signatures for drug discovery. \textit{PLoS computational biology}, 18(2): e1009909.
    \item Bharadhwaj, V. S., \textbf{Mubeen, S.}, Sargsyan, A., Jose, G. M., Geissler, S.,  … Kodamullil A.T. (2022). Integrative analysis to identify shared mechanisms between schizophrenia and bipolar disorder and their comorbidities. \textit{bioRxiv}, 2022.03.07.483233.
    \item \textbf{Mubeen, S.}, Domingo-Fern\'{a}ndez D., Díaz del Ser S., Solanki, D., Kodamullil A.T., Hofmann-Apitius M., Hopp M.T.,\textsuperscript{\textdagger} and Imhof D.\textsuperscript{\textdagger} (2022). Exploring the complex network of heme-triggered effects on the blood coagulation system. \textit{Manuscript in preparation}.


\end{enumerate}